---
title: "Intro"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(PerformanceAnalytics)
library(performance)
library(reshape2)
library(ggforce)
library(MASS)
library(car)
library(mgcv)
set.seed(123)
```

## Exploratory Data Analysis

The goal of our ananlysis is to study how the value of totale_ospedalizzati (our variable of interest) changed overtime due to other factors and to predict how it will change at the begininning of February. 

The dataset of interest contains information about the Covid-19 spreading outbreak (on 30 variables) and is taken from the official website of Protezione Civile.

To region to be analysed is Lomabrdia and the period of interest is from 1st October 2020 to 1st February 2021. We have extracted those data from the official database.

Our varibale of interest is totale_ospedalizzati.

```{r echo=FALSE}
df<-read.csv("https://raw.githubusercontent.com/pcm-dpc/COVID-19/master/dati-regioni/dpc-covid19-ita-regioni.csv",na.strings=c("",NA))

beg<-"2020-09-25T00:00:00"
end<-"2021-02-18T00:00:00"

df_sub <- df[df$denominazione_regione == 'Lombardia',]
df_sub <- df_sub[df_sub$data >= beg & df_sub$data <=end,]
summary(df_sub)

```


First, we give a brief overview to the data contained in the database. As it can be noticed some of the 30 variables miss information for around 50% or more of the observations, such as:

* `totale_positivi_test_molecolare`
* `totale_positivi_test_antigenico_rapido`
* `codice_nuts_1`
* `codice_nuts_2`
* `tamponi_test_molecolare`
* `tamponi_test_antigenico_rapido`
* `casi_da_sospetto_diagnostico`
* `casi_da_screening`
* `ingressi_terapia_intensiva`

Indeed some of these variables have been recorded olnly after the 2021-01-15. Other variables are completely missing, such as:
* `note`
* `note_test`       
* `note_casi`

We have decided to remove from the dataset that we will use for the analysis all these variables and the variables related to the region identification (such as stato, codice_regione, denominazione_regione, lat, lon).

**variables left for modeling are 11**:

* `ricoverati_con_sintomi`: Hospitalised patients with symptoms.
* `terapia_intensiva`: Intensive Care.
* `isolamento_domiciliare`: Home confinement
* `totale_positivi`: Total amount of current positive cases (Hospitalised patients + Home confinement).
* `variazione_totale_positivi`: New amount of current positive cases (totale_positivi current day - totale_positivi previous day).
* `nuovi_positivi`: New amount of current positive cases (totale_casi current day - totale_casi previous day).
* `dimessi_guariti`: Recovered.
* `deceduti`: Death (cumulated values).
* `totale_casi`: Total amount of positive cases.
* `tamponi`: Tests performed.
* `casi_testati`: Total number of people tested.

5 variables out of the former list are cumulative so they hve been converted in **per_day** data. These variables are:

* `dimessi_guariti`
* `deceduti`
* `totale_casi`
* `tamponi`
* `casi_testati`

```{r echo=FALSE}
df_sub <- df_sub[,c("data",
                    "totale_ospedalizzati",
                    "ricoverati_con_sintomi", 
                    "totale_positivi", 
                    "nuovi_positivi", 
                    "dimessi_guariti", 
                    "deceduti",
                    "variazione_totale_positivi", 
                    "terapia_intensiva",
                    "casi_testati", 
                    "totale_casi", 
                    "tamponi",
                    "isolamento_domiciliare"
)]



dimessi_guariti_per_day <- diff(df_sub$dimessi_guariti)
df_sub$dimessi_guariti_per_day <- c(dimessi_guariti_per_day[1],dimessi_guariti_per_day)

deceduti_per_day <- diff(df_sub$deceduti)
df_sub$deceduti_per_day <- c(deceduti_per_day[1],deceduti_per_day)

casi_testati_per_day <- diff(df_sub$casi_testati)
df_sub$casi_testati_per_day <- c(casi_testati_per_day[1],casi_testati_per_day)

totale_casi_per_day <- diff(df_sub$totale_casi)
df_sub$totale_casi_per_day <- c(totale_casi_per_day[1],totale_casi_per_day)

tamponi_per_day <- diff(df_sub$tamponi)
df_sub$tamponi_per_day <- c(tamponi_per_day[1],tamponi_per_day)

```


Once we have the complete dataset we have decided to insert two other variables: zone colors and rt_poitivi.
Assuming that effects of restriction rules can be seen only after a certain amount of time, we have shifted the colors by 14 days forward.

```{r, echo=FALSE}
aree_zone <- read.csv("https://raw.githubusercontent.com/ondata/covid19italia/master/webservices/COVID-19Aree/processing/areeStorico_wide.csv")
aree_zone <- aree_zone[aree_zone["datasetIniISO"] >= "2020-09-25", ]
aree_zone <- aree_zone[aree_zone["datasetIniISO"] < "2021-02-18", ]
aree_zone <- data.frame(aree_zone$datasetIniISO, aree_zone$ITC4)

column_new_dates <- data.frame(seq(from = as.Date(beg), to = as.Date(end), by= "day"))
colnames(column_new_dates) <- "data"
colnames(aree_zone) <- c("data", "color")
aree_zone$data <- as.Date(aree_zone$data)

#Remove rows with empty color
aree_zone <- aree_zone[!(is.na(aree_zone$color) | aree_zone$color==""), ]

aree_zone_merged <- merge(column_new_dates, aree_zone, by = "data", all.x = TRUE)

aree_zone_merged[1,2] <- "bianca"

last_color = "bianca"
for (i in 1:nrow(aree_zone_merged)){
    if(is.na(aree_zone_merged[i,2])){
        aree_zone_merged[i,2] = last_color
    }else{
        last_color = aree_zone_merged[i,2]
    }
}

# Delay the colors by 2 weeks: shift forward the color vector by 14 cells
# prepending 14 cells (days) of zona bianca
d_color <- aree_zone_merged$color
color_delayed <- c(rep("bianca",times=14), d_color[1:(length(d_color)-14)])
aree_zone_merged_ext <- aree_zone_merged
str(aree_zone_merged_ext)
aree_zone_merged_ext$color_delayed <- as.factor(color_delayed)

df_sub$data <- as.Date(df_sub$data)

df_sub <- merge(df_sub, aree_zone_merged_ext, by = "data")
#df_sub
df_sub$color <- as.factor(df_sub$color)
df_sub$color_delayed <- as.factor(df_sub$color_delayed)
```

```{r, echo=FALSE}
rt_lomb <- read.csv("iss_rt_lombardia.csv")
rt_lomb <- rt_lomb[rt_lomb["data"] >= "2020-09-25", ]
rt_lomb <- rt_lomb[rt_lomb["data"] < "2021-02-18", ]


rt <- data.frame(data=as.Date(rt_lomb$data), rt=rt_lomb$rt_positivi)
summary(rt)

df_sub$data<-as.Date(df_sub$data)
df_sub <- merge(df_sub, rt, by='data')

```


```{r}

df_sub$dummy = 0
i=1

for(data in df_sub$data){
    if(data >= as.Date("2020-11-22")){
        df_sub[i, "dummy" ] = 1
    }
    i = i+1
}

df_sub_ext<-df_sub

```

```{r}
inizio_train<-"2020-10-01T00:00:00"
end_train<-"2021-02-01T00:00:00"
df_sub <- df_sub[df_sub$data >= inizio_train & df_sub$data <=end_train,]
```

After having filtered the dataset, we remains with 124 observations (from 1st October 2020 to 1st February 2021) and 22 variables.

Here we plot our varible of interest over the period that we analyse:

```{r , echo=FALSE}
start_date <- as.Date("2020-10-02")
end_date <- as.Date("2021-01-31")
ggplot(df_sub, aes(x=as.Date(data), y=totale_ospedalizzati)) +
  geom_rect(data = df_sub, 
            aes(xmin=data, xmax = end_date, 
                ymin = -Inf, ymax = Inf, 
                fill = color), show.legend = FALSE) +
    #scale_x_date(date_breaks = "10 years", date_labels = "%Y", 
                #limits = c(min(as.Date(df_sub$data)), max(as.Date(df_sub$data)))) +
    scale_fill_manual(values = c("arancione" = "orange1", "gialla" = "yellow1", "bianca" = "white", "rossa" = "tomato1")) +
  geom_linerange(aes(ymax=totale_ospedalizzati, 
                     ymin=df_sub$totale_ospedalizzati[which.min(df_sub$totale_ospedalizzati)]), 
                 data=df_sub[which.max(df_sub$totale_ospedalizzati),], 
                 col="blue", lty=2) + 
  geom_text(aes(label=totale_ospedalizzati), 
            col="blue", data=df_sub[which.max(df_sub$totale_ospedalizzati),], 
            hjust=7, vjust=-0.4) + 
  geom_text(label=as.Date(df_sub[which.max(df_sub$totale_ospedalizzati),c("data")]), 
            col="blue",
            x=as.Date(df_sub$data[which.max(df_sub$totale_ospedalizzati)]),
            y=100,
            size=3) +
  geom_segment(x = as.Date(df_sub$data[1]),
               y = df_sub$totale_ospedalizzati[which.max(df_sub$totale_ospedalizzati)], 
               xend = as.Date(df_sub$data[which.max(df_sub$totale_ospedalizzati)]), 
               yend = df_sub$totale_ospedalizzati[which.max(df_sub$totale_ospedalizzati)], 
               colour = "blue",
               lty=2
               )+
  geom_line()+
  xlab("Date") +
  scale_y_continuous(breaks = seq(0,9500,by=1000))+
  theme_classic()

  
```


As it can be seen from the graph, the peak of totale_ospedalizzati has been reached the 2020-11-22 with a value of 9340 people ospedalized.
It can also be noticed that before the peak the variable of interest follows an exponential growth, while after the peak it follows a gaussian growth model. 
For this reason we have created a dummy varible to indicate if the observation is before the peak (assigning value 0) or after the peak (assigning value 1).

Here we plot the relevant varibales overtime:

```{r echo=FALSE}
df.m <- melt(df_sub[,-c(19, 20)], "data")

ggplot(df.m, aes(as.Date(data), value, group = 1)) + 
  geom_line() + 
  facet_wrap(~variable, scales = "free") +
  xlab("Date")

```

We have noticed that: `totale_ospedalizzati` is exact the sum of `ricoverati_con_sintomi` and `terapia_intensiva`, as it is shown in the graph:

```{r , echo=FALSE}
ggplot(df_sub, aes(x=as.Date(data), y=totale_ospedalizzati, col='totale_ospedalizzati')) +
  geom_line() +
  xlab("Date") +
  scale_y_continuous(breaks = seq(0,9500,by=1000))+
  geom_line(aes(x=as.Date(data), y=ricoverati_con_sintomi, col='ricoverati_con_sintomi'))+
  geom_line(aes(x=as.Date(data), y=terapia_intensiva, col='terapia_intensiva'))
  
```


Second: `totale_ospedalizzati` is the difference of: `totale_positivi` - `isolamento_domiciliare`.

```{r , echo=FALSE}
ggplot(df_sub, aes(x=as.Date(data), y=totale_ospedalizzati, col='totale_ospedalizzati')) +
  geom_line() +
  xlab("Date") +
  scale_y_continuous(breaks = seq(0,9500,by=1000))+
  geom_line(aes(x=as.Date(data), y=totale_positivi, col='totale_positivi'))+
  geom_line(aes(x=as.Date(data), y=isolamento_domiciliare, col='isolamento_domiciliare'))
  
```

Thus these 4 variables out of the list of 11 variables will be omitted. (le togliamo perche non ha senso fare dei modelli cosi?)


The remaining variables are 7:

* `variazione_totale_positivi`: not interesting. (perche???)
* `nuovi_positivi`: interesting.
* `dimessi_guariti_per_day`: interesting ?
* `deceduti_per_day`: interesting ?
* `totale_casi_per_day`: interesting ?
* `tamponi_per_day`: not interesting ?
* `casi_testati_per_day`: not interesting ?


We have also noticed that 'nuovi_positivi' is the same variable than 'totale_casi_per_day' (a varibale that we calculated). Thus we have decided to keep only 'nuovi_positivi'.

```{r , echo=FALSE}
ggplot(df_sub, aes(x=as.Date(data), y=nuovi_positivi, color="nuovi_positivi")) +
  geom_line() + 
  geom_line(data=df_sub,aes(x=as.Date(data),y=totale_casi_per_day, color="totale_casi_per_day")) +
  xlab("Date") + 
  ylab("Daily count") + 
  scale_color_manual(name="Variable", values = c(1:2))
  #scale_y_continuous(breaks = seq(0,9500,by=1000))+

```

Then we look to the scatterplot and to the corrations between the variables to try to figure out which are the most significant variables:

```{r , echo=FALSE}
chart.Correlation(df_sub[,c("nuovi_positivi", 
                            "dimessi_guariti_per_day",
                            "deceduti_per_day",
                            "casi_testati_per_day",                            
                            "tamponi_per_day", 
                            "totale_ospedalizzati"
                            )])

```

As it can be noticed they "nuovi_positivi", "dimessi_guariti_per_day", "deceduti_per_day", "tamponi_per_day" have a strong correlation with the varibale of interest. 


Now we look at the boxplots to spot outliers:

```{r , echo=FALSE, fig.width=6, fig.height=6}
# ggplot(df_sub, aes(y=nuovi_positivi)) + 
#   geom_boxplot()
varnames <- c("totale_ospedalizzati", "nuovi_positivi", "variazione_totale_positivi", 
              "dimessi_guariti_per_day", "deceduti_per_day", "casi_testati_per_day",
               "tamponi_per_day")
indexes <- c(2,5,8,14,15,16,18)
# boxplot(df_sub$nuovi_positivi,
#         xlab=varnames[2],
#         ylab="Count"
#         )
par(mfrow = c(3, 3))
invisible(lapply(indexes, function(i) boxplot(df_sub[, i], ylab="Count", main=colnames(df_sub)[i])))
#stripchart(df_sub$nuovi_positivi, method = "jitter", pch = 7, add = TRUE, col = "blue")
```

The variables with more outliers are (they have values that are external to the maximum and the minimum):

* `variazione_totale_positivi`
* `dimessi_guariti_per_day`: **GRAVE perché superano gli ospedalizzati** -> dati aggiornati con discontinuità.
* `nuovi_positivi`
* `casi_testati_per_day`: **NON GRAVE** -< sono sempre molto meno degli ospedalizzati. only one.
* `deceduti_per_day`: just two.

We will not consider `variazione_totale_positivi` anymore (perche???)

Here we plot the different varibles overtime.

```{r , echo=FALSE}
ggplot(df_sub, aes(x=as.Date(data), y=nuovi_positivi, color="nuovi_positivi")) +
  geom_line() + 
  geom_line(data=df_sub,aes(x=as.Date(data),y=totale_ospedalizzati, color="totale_ospedalizzati")) +
  geom_line(data=df_sub,aes(x=as.Date(data),y=dimessi_guariti_per_day, color="dimessi_guariti_per_day")) +
  geom_line(data=df_sub,aes(x=as.Date(data),y=deceduti_per_day, color="deceduti_per_day")) +
  geom_line(data=df_sub,aes(x=as.Date(data),y=casi_testati_per_day, color="casi_testati_per_day")) +
  geom_line(data=df_sub,aes(x=as.Date(data),y=tamponi_per_day, color="tamponi_per_day")) +
  xlab("Date") + 
  ylab("Daily count") + 
  scale_color_manual(name="Variable", values = c(1:6))
  #scale_y_continuous(breaks = seq(0,9500,by=1000))+

```

As it can be noticed from this graph and from the boxplot above 'dimessi_guariti_per_day' and 'nuovi_positivi' have a lot of outliers. 
Moreover, we look to the mean of the nubmer of 'dimessi_guariti_per_day' and 'nuovi_positivi' per day of the week. As it can be noticed they do not have a uniform distribution between the diferent days of the week, indicating that the outlier can be also due to the day in which the data are recordered.
For example: the average number of 'nuovi_positivi' is less on monday due to the fact than on sunday less tamponi are made. Insteadm the average number of 'dimessi_guariti_per_day' is less in the weekend as expected. 
```{r , echo=FALSE}

df_sub <- add_column(df_sub, weekdays(df_sub$data), .after="data")
names(df_sub)[names(df_sub) == "weekdays(df_sub$data)"] <- "giorno_settimana"

distinct_days <- as.factor(df_sub$giorno_settimana)


count_week_day_dimessi <- tapply(df_sub$dimessi_guariti_per_day, df_sub$giorno_settimana, mean, simplify = FALSE)
count_week_day_positivi <- tapply(df_sub$nuovi_positivi, df_sub$giorno_settimana, mean, simplify = FALSE)


barplot(height = as.numeric(count_week_day_dimessi), names=levels(distinct_days))

barplot(height = as.numeric(count_week_day_positivi), names=levels(distinct_days))


```

To solve the problems of the outliers we created a "moving average" for the variables 'dimessi_guariti_per_day' and 'nuovi_positivi', considering three days before and after the observations.
the two graphs represent the moving mean.

```{r, echo=FALSE}

compute_mean_points <- function(var, from, to){
    var_mean <- c()
    for (i in 1:length(var)){
        if(i >= length(var)-to){
            var_mean <- append(var_mean, mean(var[(i-from):(length(var)-((length(var))-i))])) #mean tra i valori finali
        }else if(i <= from){
            var_mean <- append(var_mean, mean(var[1:(i+to)])) #mean tra i valori iniziali
        }else{
            var_mean <- append(var_mean, mean(var[(i-from):(i+to)])) #mean nei valori di mezzo
        }

    }
    return(var_mean)
}

#Dismessi guariti per day
var_mean <- compute_mean_points(df_sub_ext$dimessi_guariti_per_day, 3, 3)
df_sub_ext$mean_dimessi_guariti_per_day <- var_mean




#Nuovi positivi
var_mean <- compute_mean_points(df_sub_ext$nuovi_positivi, 3, 3)
df_sub_ext$mean_nuovi_positivi <- var_mean


df_sub <- df_sub_ext[df_sub_ext$data >= inizio_train & df_sub_ext$data <= end_train,]

ggplot(df_sub) +
  geom_line(aes(x=as.Date(data),y=mean_dimessi_guariti_per_day, color="Moving mean")) +
  xlab("Date") + 
  ylab("Dismessi guariti") +
  geom_point(aes(x=as.Date(data),y=dimessi_guariti_per_day))

ggplot(df_sub) +
  geom_line(aes(x=as.Date(data),y=mean_nuovi_positivi, color="Moving mean")) +
  xlab("Date") + 
  ylab("Nuovi positivi") +
  geom_point(aes(x=as.Date(data),y=nuovi_positivi))



```

The new variables overtime are:

```{r , echo=FALSE}
ggplot(df_sub, aes(x=as.Date(data), y=mean_nuovi_positivi, color="mean_nuovi_positivi")) +
  geom_line() + 
  geom_line(data=df_sub,aes(x=as.Date(data),y=totale_ospedalizzati, color="totale_ospedalizzati")) +
  geom_line(data=df_sub,aes(x=as.Date(data),y=mean_dimessi_guariti_per_day, color="mean_dimessi_guariti_per_day")) +
  geom_line(data=df_sub,aes(x=as.Date(data),y=deceduti_per_day, color="deceduti_per_day")) +
  geom_line(data=df_sub,aes(x=as.Date(data),y=mean_dimessi_guariti_per_day*dummy, color="mean_dimessi_guariti_per_day*dummy")) +
  #geom_line(data=df_sub,aes(x=as.Date(data),y=casi_testati_per_day, color="casi_testati_per_day")) +
  #geom_line(data=df_sub,aes(x=as.Date(data),y=tamponi_per_day, color="tamponi_per_day")) +
  xlab("Date") + 
  ylab("Daily count") + 
  scale_color_manual(name="Variable", values = c(1:5))
  #scale_y_continuous(breaks = seq(0,9500,by=1000))+

```

Here we show the scatterplots between the variable of interest and the other variables (before and after the moving mean):

```{r, echo=FALSE}

data_max_nuovi_positivi=df_sub[which.max(df_sub$mean_nuovi_positivi),]
nuovi_positivi_max_split1 <- df_sub[df_sub$data < data_max_nuovi_positivi$data, ]
nuovi_positivi_max_split2 <- df_sub[df_sub$data >= data_max_nuovi_positivi$data, ]

plot1 <- ggplot() +
  geom_point(aes(x=df_sub$totale_ospedalizzati[df_sub$data < data_max_nuovi_positivi$data], y=nuovi_positivi_max_split1$nuovi_positivi), color="red") + 
  geom_point(aes(x=df_sub$totale_ospedalizzati[df_sub$data >= data_max_nuovi_positivi$data], y=nuovi_positivi_max_split2$nuovi_positivi), color="orange") + 
  xlab("totale_ospedalizzati") + 
  ylab("nuovi_positivi")

plot2 <- ggplot(df_sub, aes(x=totale_ospedalizzati, y=dimessi_guariti_per_day)) +
  geom_point(color="green") + 
  xlab("totale_ospedalizzati") + 
  ylab("dimessi_guariti_per_day")

plot3 <- ggplot(df_sub, aes(x=totale_ospedalizzati, y=deceduti_per_day)) +
  geom_point(color="black") + 
  xlab("totale_ospedalizzati") + 
  ylab("deceduti_per_day")

data_max_nuovi_positivi=df_sub[which.max(df_sub$mean_nuovi_positivi),]
nuovi_positivi_max_split1 <- df_sub[df_sub$data < data_max_nuovi_positivi$data, ]
nuovi_positivi_max_split2 <- df_sub[df_sub$data >= data_max_nuovi_positivi$data, ]

plot4 <- ggplot() +
  geom_point(aes(x=df_sub$totale_ospedalizzati[df_sub$data < data_max_nuovi_positivi$data], y=nuovi_positivi_max_split1$mean_nuovi_positivi), color="red") + 
  geom_point(aes(x=df_sub$totale_ospedalizzati[df_sub$data >= data_max_nuovi_positivi$data], y=nuovi_positivi_max_split2$mean_nuovi_positivi), color="orange") + 
  xlab("totale_ospedalizzati") + 
  ylab("mean_nuovi_positivi")

plot5 <- ggplot(df_sub, aes(x=totale_ospedalizzati, y=mean_dimessi_guariti_per_day)) +
  geom_point(color="green") + 
  xlab("totale_ospedalizzati") + 
  ylab("mean_dimessi_guariti_per_day")
library(ggpubr)
ggarrange(plot1, plot2, plot3, plot4, plot5, ncol = 3, nrow = 3)
```

deceduti_per_day has a linear relationship with totale_ospedalizzati while mean_dimmessi_guariti_per_day seems to have a square relation with respect to totale_ospedalizzati. (e mean_nuovi_positivi_per day is almost linear until the peak of mean_nuovi_positivi).




## Linear models 



















